---
title: "GEO-TREES tree inventory QA/QC"
author:
  - "John L. Godlee"
  - "Vincyane Badouard"
date: "`r Sys.Date()`"
version: "0.1"
format: 
  html:
    embed-resources: true
    toc: true
    toc-location: left
    toc-title: "Contents"
    number-sections: true
    code-fold: "show"
    df-print: "paged"
    page-layout: full
    html-math-method: katex
    other-links:
      - text: GEO-TREES website
        href: https://geo-trees.org
    code-links:
      - text: GitHub
        href: https://github.com/GEO-TREES/INV_quality_control
jupyter: ir
engine: knitr
---

Version: **{{< meta version >}}**

# Introduction

This interactive R notebook provides a workflow for quality assurance and
quality control (QA/QC) of GEO-TREES tree inventory data. 

This notebook is provided as both a `.qmd` Quarto file, which we recommend
running in RStudio, and as a `.ipynb` Jupyter Notebook, which you can run in
JupyterLab either locally or on the ESA/NASA MAAP platforms. You can use either
version to QA/QC your data.

The process for QA/QC:

1. Create a copy of your raw stem measurement data that conforms to the
   GEO-TREES L0 data format. See the [GEO-TREES Tree Inventory Guidelines, _Section 16: Data format_](https://docs.google.com/document/d/1Q5Wy_LbNEa0tNaGVntL0Bku41Z3t-TeU85_IxZuC6Uc/edit?usp=sharing). 
   You can use the [ATFS Data Harmonization Shiny app](https://docs.google.com/document/d/1Q5Wy_LbNEa0tNaGVntL0Bku41Z3t-TeU85_IxZuC6Uc/edit?usp=sharing)
   to help with this if you wish. The `inst/` folder in [this repository](https://github.com/GEO-TREES/INV_quality_control) 
   contains a `.rds` output profile for the GEO-TREES data format. __The
   original raw data should not be altered__.
2. Open the notebook and load your __copy__ of the L0 tree inventory data.
3. Run the code chunks interactively in the notebook. Look at the outputs, and
   make changes to the input data where necessary. Changes can be made either
   by adding code to the notebook, or manually editing your __copy__ of the L0
   tree inventory data. If you choose to manually edit your tree inventory
   data, make a second copy of the data for this purpose.
4. Once you have finished working through the notebook, save and render the
   notebook as a HTML file. Note that all code in the notebook must run without
   errors for the file to render successfully. The method depends on whether
   you are using the `.qmd` or `.ipynb` version:
    * `.qmd` in RStudio: Run `quarto::quarto_render("PDA_QAQC_nb.qmd",
      output_format = "html")` in the R console. You may need to run
      `install.packages("quarto")` first.
    * `.ipynb` in JupyterLab: Click `File > Save and Export Notebook As > HTML`
5. The following data should be sent to GEO-TREES for data archiving:
    * Original version of the raw data
    * Formatted raw data in the GEO-TREES L0 data format.
    * Raw data with changes made following QA/QC
    * The completed `.qmd` / `.ipynb` QA/QC notebook.
    * Rendered HTML of the QA/QC notebook 

# Setup

Install and load the GEO-TREES tree inventory QA/QC R package contained in this repository. This package
contains various convenience functions for flagging and optionally correcting
tree inventory data. The package will also load various other packages as
dependencies. If you want to load additional packages to help you explore and
manipulate your tree inventory data, you can add to the code chunk below.


```{r}
#| eval: false
devtools::install_github("GEO-TREES/INV_quality_control")
```


```{r}
library(INV_quality_control)
library(BIOMASS)  # Match taxonomic names to WFO `correctTaxo()`, fit diameter-height models `modelHD()`
library(ggplot2)  # Augment `plotScatter()` and `plotHistogram()` 
library(tibble)  # `tribble()` - create dataframes (tibbles) by row
library(DT)  # Pretty tables in HTML notebooks
```

```{r}
#| echo: false
# Change default displayHelp rendering if compiled as HTML
if (getOption("knitr.in.progress")) {
  formals(displayHelp)$html <- TRUE
} 
```

Throughout the notebook we display the help files for the different functions
in the notebook using the `displayHelp()` function. Below is the help file for
the `INV_quality_control` R package:

```{r}
#| echo: false
displayHelp(INV_quality_control)
```

# Create L0 tree inventory data   

Create a __copy__ of your raw stem measurement data that conforms to the GEO-TREES 
L0 data format. See the [GEO-TREES Tree Inventory Guidelines, _Section 16: Data format_](https://docs.google.com/document/d/1Q5Wy_LbNEa0tNaGVntL0Bku41Z3t-TeU85_IxZuC6Uc/edit?usp=sharing). 
__The original raw data should not be altered, always work with a copy of the data__.

You can use the [ATFS Data Harmonization Shiny app](https://docs.google.com/document/d/1Q5Wy_LbNEa0tNaGVntL0Bku41Z3t-TeU85_IxZuC6Uc/edit?usp=sharing) 
to help re-format your stem measurement data if you wish. The `inst/` folder in
[this repository](https://github.com/GEO-TREES/INV_quality_control) 
contains a `.rds` output profile for the GEO-TREES data format, which you can
load into the app. The app will walk you through creating an input profile for
your data, then performs the re-formatting procedure.

Alternatively, you can manually re-format your data, for example using a
spreadsheet editor, or in R.

The QA/QC code below requires you to supply the stem measurement table, the
plot meta-data table, the plot GNSS table. 

You can also optionally provide the taxonomy table, stem height measurement
table, and wood density measurement table.

# Load data

Edit the code below to load your tree inventory data in the GEO-TREES L0 data
format.

Load stem measurement table:

```{r}
s <- read.csv("../../bicuar_test_data/out/stem.csv")
```

Load plot meta-data table:

```{r}
p <- read.csv("../../bicuar_test_data/out/plot.csv")
```

Load plot geo-location table:

```{r}
pt <- read.csv("../../bicuar_test_data/out/plot_pt.csv")
```

Optionally load taxonomy table:

```{r}
#| eval: false
tx <- read.csv("taxon_example.csv")
```

Optionally load additional stem height data:

```{r}
#| eval: false
h <- read.csv("height_example.csv")
```

Optionally load additional wood density data:

```{r}
#| eval: false
wd <- read.csv("wood_density_example.csv")
```

# Missing, duplicated, and phantom records

## Missing records

```{r}
#| echo: false
displayHelp(flagRecordMissing)
```

```{r}
df_flagRecordMissing <- flagRecordMissing(
  x = s,
  ind_id = c("site_id", "plot_id", "tree_id", "stem_id"),
  census_id = "census_id",
  census_id_vec = c(1, 2, 3),  # Vector of census IDs
  comment = "Missing stem measurement record"
)

datatable(df_flagRecordMissing)
```

Missing records should be added to the stem measurement table. Missing
measurement values may be estimated by e.g. `correctDiameterSeries()` and
`correctTaxonSeries()`.

The `datatable()` function above produces pretty HTML tables in a Quarto
document.

## Duplicated records

```{r}
#| echo: false
displayHelp(flagRecordDup)
```

```{r}
# Flag duplicated records based on ID values
df_flagRecordDup <- flagRecordDup(
  x = s, 
  meas_id = c("site_id", "plot_id", "tree_id", "stem_id", "measurement_id"),
  census_id = "census_id"
)

datatable(df_flagRecordDup)
```

Duplicated records should be merged or removed. There should only be one
measurement for each concatenation of
`site_id`+`plot_id`+`tree_id`+`stem_id`+`measurement_id`.

## Missing values

```{r}
#| echo: false
displayHelp(flagValMissing)
```

```{r}
df_flagValMissing <- flagValMissing(
  x = s,
  meas_id = c("site_id", "plot_id", "tree_id", "stem_id", "measurement_id"),
  target = c(
    "site_id", 
    "plot_id", 
    "tree_id", 
    "stem_id", 
    "census_id", 
    "measurement_date", 
    "taxon_name", 
    "x_rel_m", 
    "y_rel_m"
  ),
  cond = NULL,
  comment = "Missing measurement values"
)

datatable(df_flagValMissing)
```

Some columns must never contain `<NA>` values, particularly ID values. 

```{r}
df_flagValMissing_diam_cm <- flagValMissing(
  x = s,
  meas_id = c("site_id", "plot_id", "tree_id", "stem_id", "measurement_id"),
  target = "diam_cm",
  cond = grepl("A", s$code) & !grepl("T|F", s$code),
  comment = "Missing diameter values"
)

datatable(df_flagValMissing_diam_cm)
```

Some columns are only allowed to contain missing values under certain
conditions. For example, living and standing stems should always have a
diameter measurement.

```{r}
df_flagValMissing_pom_m <- flagValMissing(
  x = s,
  meas_id = c("site_id", "plot_id", "tree_id", "stem_id", "measurement_id"),
  target = "pom_m",
  cond = !is.na(s$diam_cm),
  comment = "Missing POM values"
)

datatable(df_flagValMissing_pom_m)
```

If a stem has a diameter measurement then it should have a Point of Measurement (POM) value.

# Outlier diameter and height measurements

```{r}
#| echo: false
displayHelp(plotHistogram)
```

Inspect the distribution of diameter measurements. Are all measurements larger
than the minimum diameter threshold of the plot? Are any measurements
unrealistically large? 

```{r}
# Plot a histogram
plotHistogram(
	x = s$diam_cm, 
	boundary = 0,  # Align bars with 0
	colour = "black",  # Make pretty 
	fill = "grey"
) + 
  xlab("Stem diameter (cm)")  # Add x axis label

# Summarise values
summary(s$diam_cm)
```

If height measurements are available, find potentially erroneous measurements
based on the relationship between height and diameter measurements.

```{r}
#| echo: false
displayHelp(flagModOutlier)
```

```{r}
# Fit a simple height-diameter model using the BIOMASS package
hmod <- BIOMASS::modelHD(
  D = s$diam_cm, 
  H = s$height_m,
  method = "weibull",
  bayesian = FALSE,
  drawGraph = TRUE)

# Flag outlier measurements using model residuals
df_flagModOutlier_height_m_diam_cm <- flagModOutlier(
  x = s, 
  meas_id = c("site_id", "plot_id", "tree_id", "stem_id", "measurement_id"),
  mod = hmod$model, 
  x_var = "diam_cm",
  y_var = "height_m",
  threshold = 3,
  comment = "Height-diameter model outlier"
)

datatable(df_flagModOutlier_height_m_diam_cm)
```

```{r}
#| echo: false
displayHelp(plotScatter)
```

```{r}
# Create scatter plot with outliers flagged
s_hd <- merge(s, df_flagModOutlier_height_m_diam_cm, all.x = TRUE)

plotScatter(
  x = s_hd$diam_cm, 
  y = s_hd$height_m, 
  flag = s_hd$comment,
  alpha = 0.5)
```

These outliers are not necessarily errors, but do deserve closer inspection.

```{r}
#| echo: false
displayHelp(flagValRange)
```

```{r}
df_flagValOutlier_diam_cm <- flagValRange(
  x = s,
  meas_id = c("site_id", "plot_id", "tree_id", "stem_id", "measurement_id"),
  target = "diam_cm",
  threshold = c(0, 200),
  comment = "Diameter measurements outside range"
)

datatable(df_flagValOutlier_diam_cm)
```

```{r}
df_flagValOutlier_height_m <- flagValRange(
  x = s,
  meas_id = c("site_id", "plot_id", "tree_id", "stem_id", "measurement_id"),
  target = "height_m",
  threshold = c(0, 50),
  comment = "Height measurements outside range"
)

datatable(df_flagValOutlier_height_m)
```

## Unit mixtures

```{r}
#| echo: false
displayHelp(flagValUnit)
```

```{r}
df_flagValUnit <- flagValUnit(
  x = s,
  plot_id = "plot_id",
  census_id = "census_id",
  target = "diam_cm",
  threshold = 5, 
  cv_threshold = 100,
  comment = "Diameter unit-shift"
)

datatable(df_flagValUnit)
```

## Rounded values

```{r}
#| echo: false
displayHelp(flagValRound)
```

```{r}
df_flagValRound_diam_cm <- flagValRound(
  x = s, 
  census_id = c("plot_id", "census_id"),
  target = "diam_cm",
  precision = 1, 
  threshold = 0.3,
  comment = "Rounded diameter values"
)

datatable(df_flagValRound_diam_cm)
```

`flagValRound()` finds groups with improbably high proportions of rounded
measurement values, which might imply poor measurement precision.

# Stem condition codes

```{r}
#| echo: false
displayHelp(flagCodeVal)
```

```{r}
df_flagCodeVal <- flagCodeVal(
  x = s,
  meas_id = c("site_id", "plot_id", "tree_id", "stem_id", "measurement_id"),
  target = "code",
  accepted = c("A", "D", "S", "F", "M", "B", "T"),
  invalid_comb = list(
	c("A", "D"),
	c("S", "F"),
	c("B", "T"),
	c("M", "A"),
	c("M", "D"),
	c("M", "S"),
	c("M", "F"),
	c("M", "B"),
	c("M", "T")
  ),
  cond = NULL,
  comment = "Invalid code values"
)

datatable(df_flagCodeVal)
```

`flagCodeVal()` identifies measurements either with invalid values in the `code` column, or invalid combinations or codes. For example, a stem cannot be dead ("D") and alive ("A") at the same time, or standing ("S") and fallen ("F").

# Stem coordinate issues

## Stem coordinate outliers

```{r}
#| echo: false
displayHelp(plotScatter)
```

Inspect the spatial distribution of stems within each plot. Are any stems
outside the plot, or distributed unevenly within the plot?

```{r}
# Plot stem coordinates
plotScatter(s$x_rel_m, s$y_rel_m) + 
  coord_equal() # +
  # facet_wrap(~s$plot_id)
```

```{r}
df_flagValOutlier_x <- flagValRange(
  x = s,
  meas_id = c("site_id", "plot_id", "tree_id", "stem_id", "measurement_id"),
  target = "x_rel_m",
  threshold = c(0, 100),
  comment = "X coordinate outside range"
)

df_flagValOutlier_y <- flagValRange(
  x = s,
  meas_id = c("site_id", "plot_id", "tree_id", "stem_id", "measurement_id"),
  target = "y_rel_m",
  threshold = c(0, 100),
  comment = "Y coordinates outside range"
)

names(df_flagValOutlier_x)[6] <- "xy_rel_m"
names(df_flagValOutlier_y)[6] <- "xy_rel_m"

datatable(rbind(df_flagValOutlier_x, df_flagValOutlier_y))
```

## Stem coordinates relative to subplots

`correctLocSub()` checks whether XY coordinates match with subplot coordinates
and optionally re-locates incongruously located stems to the center of their
subplot.

```{r}
# Example dataframe with information on subplot relative coordinates
# Replace with example from your plots. 
subs <- tribble(
  ~subplot_id, ~x_min, ~x_max, ~y_min, ~y_max,
  1,           0,      100,    0,      10,
  2,           0,      100,    10,     20,
  3,           0,      100,    20,     30,
  4,           0,      100,    30,     40,
  5,           0,      100,    40,     50,
  6,           0,      100,    50,     60,
  7,           0,      100,    60,     70,
  8,           0,      100,    70,     80,
  9,           0,      100,    80,     90,
  10,          0,      100,    90,     100
)
```

```{r}
#| echo: false
displayHelp(correctLocSub)
```

```{r}
# If the subplot layout varies by plot you can create unique subplot IDs for
# each plot in the `subs` dataframe with multiple ID columns.

# Check if stem measurements located inside the correct subplot 
# according to relative coordinates and subplot ID.
df_correctLocSub <- correctLocSub(
  x = s,
  meas_id = c("site_id", "plot_id", "tree_id", "stem_id", "measurement_id"),
  subplot_id = "subplot_id",
  x_rel = "x_rel_m",
  y_rel = "y_rel_m",
  subs = subs,
  subs_subplot_id = "subplot_id",
  subs_x_min = "x_min",
  subs_x_max = "x_max",
  subs_y_min = "y_min",
  subs_y_max = "y_max",
  correct = TRUE,
  comment = "Stem coordinates outside subplot"
)

datatable(df_correctLocSub)
```

## Coordinates of multi-stemmed trees

```{r}
#| echo: false
displayHelp(flagMultiLoc)
```

```{r}
df_flagMultiLoc <- flagMultiLoc(
  x = s,
  ind_id = c("site_id", "plot_id", "tree_id"),
  census_id = "census_id",
  x_rel = "x_rel_m",
  y_rel = "y_rel_m",
  threshold = 5,
  comment = "Multi-stemmed tree with incongruous stem coordinates"
)

datatable(df_flagMultiLoc)
```

# Taxonomic name issues 

## Basic orthographic and nomenclature issues

```{r}
#| echo: false
displayHelp(flagTaxonName)
```

```{r}
df_flagTaxonName <- flagTaxonName(
  x = s,
  meas_id = c("site_id", "plot_id", "tree_id", "stem_id", "measurement_id"),
  taxon_name = "taxon_name",
  family = NULL, 
  genus = NULL,
  species = NULL,
  regex = "[^a-zA-Z ]",
  comment = "Taxonomic name issue"
)

datatable(df_flagTaxonName)
```

## Taxonomic names of multi-stemmed trees

```{r}
#| echo: false
displayHelp(flagMultiTaxon)
```

```{r}
df_flagMultiTaxon <- flagMultiTaxon(
  x = s,
  ind_id = c("site_id", "plot_id", "tree_id"),
  census_id = "census_id",
  taxon_name = "taxon_name",
  comment = "Multi-stemmed tree with multiple taxonomic names"
)

datatable(df_flagMultiTaxon)
```

## Check taxonomic names against the World Flora Online database

```{r}
#| echo: false
displayHelp(correctTaxo, pkg = "BIOMASS")
```

```{r}
df_correctTaxo <- BIOMASS::correctTaxo(
	s$taxon_name,
	interactive = FALSE,
	preferAccepted = TRUE,
	preferFuzzy = TRUE,
	sub_pattern = BIOMASS::subPattern(),
	useCache = FALSE,
	useAPI = TRUE,
	capacity = 60,
	fill_time_s = 60, 
	timeout = 10)

datatable(unique(df_correctTaxo[df_correctTaxo$nameModified == TRUE,]))
```

# Time-series inconsistencies

__This part of the notebook should only be used if you have access to
multi-census data.__

Even if you are only contributing one tree inventory census to GEO-TREES you
can still use data from multiple censuses to identify potential errors in the
GEO-TREES census.

## Diameter timeline

```{r}
# Create "alive" mortality status column
s$alive <- NA
s$alive[grepl("A", s$code)] <- TRUE
s$alive[grepl("D", s$code)] <- FALSE
# Note: If a code contains both A and D, the last line 
# evaluated (D) will take precedence.
```

```{r}
#| echo: false
displayHelp(correctDiameterSeries)
```

```{r}
df_correctDiameterSeries <- correctDiameterSeries(
  x = s,
  ind_id = c("site_id", "plot_id", "tree_id", "stem_id"),
  plot_id = c("site_id", "plot_id"),
  census_date = "census_date",
  diam = "diam_cm",
  status = "alive",
  pom = "pom_m",
  growth_thresh = c(-2, 4),
  n_stem_thresh = 3, 
  digits = 1,
  comment = "Diameter timelines corrected"
)

datatable(df_correctDiameterSeries)
```

## Mortality timeline

```{r}
#| echo: false
displayHelp(correctMortalitySeries)
```

```{r}
df_correctMortalitySeries <- correctMortalitySeries(
  x = s,
  ind_id = c("site_id", "plot_id", "tree_id", "stem_id"),
  census_id = "census_id",
  diam = "diam_cm",
  status = "alive",
  comment = "Mortality timelines corrected"
)

datatable(df_correctMortalitySeries)
```

## Stem condition timeline

```{r}
#| echo: false
displayHelp(flagCodeSeries)
```

```{r}
df_flagCodeSeries <- flagCodeSeries(
  x = s,
  ind_id = c("site_id", "plot_id", "tree_id", "stem_id"),
  census_id = "census_id",
  target = "code",
  invalid_trans = list(
    c("D", "A"),
    c("F", "S"),
    c("B", NA_character_),
    c("T", NA_character_),
    c("T", NA_character_)
  ),
  comment = "Invalid code transition"
)

datatable(df_flagCodeSeries)
```

## Taxonomy timeline

```{r}
#| echo: false
displayHelp(flagTaxonSeries)
```

```{r}
df_flagTaxonSeries <- flagTaxonSeries(
  x = s,
  ind_id = c("site_id", "plot_id", "tree_id", "stem_id"),
  census_id = "census_id",
  taxon_name = "taxon_name",
  comment = "Taxon name differs among censuses"
)

datatable(df_flagTaxonSeries)
```

## Overgrown recruits

```{r}
#| echo: false
displayHelp(flagRecruitBig)
```

```{r}
df_flagRecruitBig <- flagRecruitBig(
  x = s,
  ind_id = c("site_id", "plot_id", "tree_id", "stem_id"),
  diam = "diam_cm",
  census_date = "census_date",
  census_date_vec = NULL, 
  min_diam_threshold = 10,  # Minimum diameter threshold of the plot(s)
  growth_max = 5,  # Maximum feasible annual diameter growth rate 
  comment = "Overgrown recruit"
)

datatable(df_flagRecruitBig)
```

## Pre-recorded recruits

```{r}
#| echo: false
displayHelp(flagRecruitPre)
```

```{r}
df_flagRecruitPre <- flagRecruitPre(
  x = s,
  ind_id = c("site_id", "plot_id", "tree_id", "stem_id"),
  census_id = "census_id",
  target = c("diam_cm"),
  comment = "Pre-census recruit"
)

datatable(df_flagRecruitPre)
```


# Miscellaneous checks

## Too many stems on multi-stemmed tree

```{r}
#| echo: false
displayHelp(flagMultiMany)
```

```{r}
df_flagMultiMany <- flagMultiMany(
  x = s,
  ind_id = c("site_id", "plot_id", "tree_id"),
  census_id = "census_id",
  threshold = 12,
  comment = "Tree with too many stems"
)

datatable(df_flagMultiMany)
```

# Create output objects

Create a list of flag dataframes and write to file:

```{r}
flag_list <- list(
  flagRecordMissing = df_flagRecordMissing,
  flagRecordDup = df_flagRecordDup,
  flagValMissing = df_flagValMissing,
  flagValMissing_diam_cm = df_flagValMissing_diam_cm,
  flagValMissing_pom_m = df_flagValMissing_pom_m,
  flagRecruitBig = df_flagRecruitBig,
  flagRecruitPre = df_flagRecruitPre,
  flagModOutlier_height_m_diam_cm = df_flagModOutlier_height_m_diam_cm,
  flagValOutlier_diam_cm = df_flagValOutlier_diam_cm,
  flagValOutlier_height_m = df_flagValOutlier_height_m,
  flagValUnit = df_flagValUnit,
  flagValRound_diam_cm = df_flagValRound_diam_cm,
  flagCodeVal = df_flagCodeVal,
  correctLocSub = df_correctLocSub,
  flagMultiLoc = df_flagMultiLoc,
  flagTaxonName = df_flagTaxonName,
  flagMultiTaxon = df_flagMultiTaxon,
  correctTaxo = df_correctTaxo,
  correctDiameterSeries = df_correctDiameterSeries,
  correctMortalitySeries = df_correctMortalitySeries,
  flagCodeSeries = df_flagCodeSeries,
  flagTaxonSeries = df_flagTaxonSeries,
  flagMultiMany = df_flagMultiMany
)

saveRDS(flag_list, "./flag_list.rds")
```
